\stktokens{}: Enforcing Well-Bracketed Control Flow and Stack Encapsulation Using Linear Capabilities

### 2019-07-26 popl-2019-journal-v/proving-full-abstraction.tex

We call the invariants regions because, as we will see in Section~\ref{subsubsec:ft-and-revocation}, they turn out to not be invariant.
A region describes a collection of related memory segments, so it is simply represented as a relation over memory segments (i.e.\ a relation in $\Rel{\MemSeg \times \MemSeg}$).
Intuitively, we want to be able to say that two memory segments are related when their contents are related.
That is, for every address in the two memory segments, the words that reside there must be related.
In Section~\ref{subsec:logical-relation}, we define precisely what it means for words to be related, but for now we provide some the intuition.
If we have two integers, then they are related when they are equal.
If we instead have two capabilities, then they should also be related if they, in some sense, are equal.
But what does it mean for capabilities to be \textit{equal}?
Intuitively, it should mean that the capabilities give you the same authority, for instance, if you have two executable capabilities, they should observably perform the same computation.
As a capability points to a piece of memory, the authority of the capability depends on the contents of that memory.
In other words to say whether two capabilities are related, we must know the possible contents of the memory.
The world expresses the possible memory contents, so our regions are world indexed, i.e.
\[
  \Worldh = \RegName \parfun (\World \fun \Rel{\MemSeg \times \MemSeg})
\]
At this point, we can see that we have constructed a recursive domain equation.
If we inline $\Worldh$ in $\World$, then we have a circular equation with no solution because the self-reference happens in a negative position.
Luckily, we can solve circular equations if we move to a different domain.
For now, we will ignore the problem and return to the issue in Section~\ref{subsubsec:rec-dom-eq}.

For the sake of readability, we introduce the following notation
\begin{gather*}
  \pwheap = \pi_1(W)\\
  \pwpriv = \pi_2(W)\\
  \pwfree = \pi_3(W)
\end{gather*}
\subsubsection{Linearity}
% Linearity
The linear capabilities of \srccm{} and \trgcm{} guarantee that they have the sole authority over the memory they reference\footnote{Under the assumption that the system was initialised with unique linear capabilties.}.
In order to model this uniqueness, we need to keep track of which parts of memory that are uniquely referenced and make sure that only one linear capability references the unique parts.
%% Bookkeeping done by worlds
We use the world to keep track of what parts of memory must be uniquely referenced by having two kinds of regions: shared and spatial.
If a memory segment is governed by a shared region, then normal capabilities may reference it.
On the other hand, if a memory segment is governed by a spatial region, then only a linear capability may reference it.
%%% spatial/spatial_owned
We cannot let multiple linear capabilities reference the same memory, so we add ownership to spatial regions.
Only the spatial owned regions can be referenced by a capability.
Even though only spatial owned regions can be referenced by linear capabilities, we still have unclaimed spatial regions that specify what part of memory looks like.
When there is an unclaimed spatial region, then there may be a compatible world  that claims the ownership of that spatial region which means that the memory may still be referenced (we return to the notion of compatible worlds w.r.t. ownership in Section~\ref{subsubsec:joining-worlds}).
Specifically, we add tags $\spatial$ and $\spatialo$ to the spatial regions:
\[
  \Regions = \left\{
  \begin{array}{l}
    \{\spatial \} \times (\World \fun \Rel{\MemSeg \times \MemSeg}) + \\
    \{\spatialo \} \times (\World \fun \Rel{\MemSeg \times \MemSeg}) \\ 
  \end{array} \right.
\]
For readability, we also add a tag $\pure$ to the shared regions:
\[
  \Regionh = \{\pure \} \times (\World \fun \Rel{\MemSeg \times \MemSeg}) 
\]

% What parts of memory are linear
We will extend the regions further in Sections~\ref{subsubsec:seals}~and~\ref{subsubsec:ft-and-revocation}, but for now we continue the definitions of the three sub-worlds.
The sub-world $\Worldh$ specifies the heap memory which can be referenced by both linear and normal capabilities, so it should contain both shared and spatial regions.
For this reason, it is defined as
\[
  \Worldh = \RegName \parfun (\Regionh + \Regions)
\]
\srccm{} internalizes the \stktokens{} stack, so it can only be referenced by linear capabilities which means that the two stack regions should only have spatial regions.
For instance the $\Worldfs$ is defined as
\[
  \Worldfs = \RegName \parfun \Regions
\]
$\Worlds$ can also only be referenced by linear capabilities, so it should also only have spatial regions.
$\Worlds$ not only models the memory contents of the local stack frames, it also models the call stack that consists of the stack frames.
Conceptually, this means that each of the stack frames is connected with a return point in some code.
In a traditional C calling convention, the code return point would even be stored in the stack frame.
However, with \stktokens{} a stack frame does not contain any information about the corresponding code return point.
Instead, the stack frame and code return point is connected by the capabilities that reference them as they are sealed with the same seal and together then constitute the sealed return pair.
In order to ensure that each call actually returns to the correct point of the code, we must still include the address of the return point in our model.
To this end, each shared region in $\Worlds$ must be paired with a return address:
\[
\Worlds = \RegName \parfun (\Regions \times \Addr)
\]

@@

\subsubsection{Future worlds and revocation}
\label{subsubsec:ft-and-revocation}
% Memory evolves over time, we need to model this.
During machine execution, the memory contents changes and parts of memory may be repurposed.
The world must model such memory changes.
% Future World
To this end, we have a future world relation $\future$ that describes how worlds and thus memory may change over time.
In order to relate \trgcm{} and \srccm{}, we are mainly interested in changes to the stack as it repurposes stack frames.
Specifically, we need to introduce regions to govern new encapsulated stack frames, and we need to revoke regions that govern call frames that have been popped from the call stack, so the stack frame can be reused for a new call frame.
We allow introduction of new regions simply by extension.
That is, a future world can have new regions that was not present in a past world.
Things are a bit more complicated when it comes to revocation.
\lau{Can we give a more precise explanation of why regions cannot be removed from the world (still in terms of intuition).}
For technical reasons, we need the worlds to be extensional, so we cannot simply forget about regions that we want to remove.
We keep removed regions in the world and tag them as removed with a $\revoked$ tag.
This means that the regions should be able to get the $\revoked$ tag which we express in a future region relation.
Not all regions that should be revocable.
Specifically, it should only be the regions that specify memory governed by linear capabilities as linear capabilities, in some sense, can be revoked.
A linear capability is the sole authority over some memory, so when a program possesses a linear capability, it knows for sure that no one else has access to that memory which makes it safe to repurpose.
On the other hand, normal capabilities may have an alias, so it is not safe to repurpose the memory a normal capability has authority over.
Only spatial regions specify memory governed by linear capabilities, so we only add the revoked tag to the $\Regions$.
The $\revoked$ region does not specify any memory, so we drop the world indexed relation.
That is, we have
\[
  \Regions = \left\{
  \begin{array}{l}
    \{\spatial \} \times (\World \fun \Rel{\MemSeg \times \MemSeg}) + \\
    \{\spatialo \} \times (\World \fun \Rel{\MemSeg \times \MemSeg}) + \\
    \{\revoked \}
  \end{array} \right.
\]
The rules for the future region relation are displayed in Figure~\ref{fig:ft-reg-rel}.
% Revocation of spatial regions. 'revocation' of linear capabilities.
The first rule says that all regions are future regions of themselves, so regions are allowed to stay unchanged.
The second rule says that $\spatial$ regions can become $\revoked$ which means that nothing can depends on the region anymore.
The third and final rule says that $\spatial$ regions can become $\spatialo$ which indicates that the linear capabilities are affine.
Specifically, the linear capabilities in \trgcm{} and \srccm{} can be dropped by overwriting them in registers or memory which makes the linear capabilities affine.
The world a capability is safe with respect to is no longer needed when the capability is dropped, so other worlds may pick up the ownership which the future region relation allows.
% describe why "Spatial -> spatialo" ~ affine or at least why we have it?
\begin{figure}[htb]
  \centering
  \begin{mathpar}
  \inferrule{ r \in \Regions \cup \Regionh }{ r \future r}
  \and
  \inferrule{ }{ \revoked \future (\spatial,\_)}
  \and
  \inferrule{ }{ (\spatialo,H) \future (\spatial,H)}
\end{mathpar}
  \caption{Future region relation.}
  \label{fig:ft-reg-rel}
\end{figure}

With the future region relation in place, we define the future world relation as follows: For worlds $W$ and $W'$,
\[
  W' \future W \text{ iff } \left\{
    \array{l}
    \text{for $i \in \{\mathrm{heap},\mathrm{free},\mathrm{priv} \}$} \\
    \quad\exists m_i : \RegionName \fun \RegionName, \text{ injective}\ldotp\\
    \qquad \dom(W'.i) \supseteq \dom(m_i(W.i)) \wedge \forall r \in \dom(W.i)\ldotp W'.i(m_i(r)) \future W.i(r)
     \endarray
  \right.
\]
The relation says that each of the three worlds must be an extension of the past world and each of the existing region must have a future region.
% injective function
Note that the future world relation has a mapping function $m_i$ which allows us to change the naming of regions in future worlds.
The definition is a generalization of the standard definition where $m_i$ would be the identity\footnote{In \citet{skorstengaard_reasoning_2017} the future region relation and the reasoning about the awkward example could have been simplified with this future world relation.}.

The memory invariants are world dependent, but the worlds may change according to the future world relation, so what does that mean for the memory invariants?
The memory invariants describe what certain parts of memory looks like.
In future worlds, part of the memory may change but for the memory invariant to be invariant, it should keep relating the same memory segments even when the rest of the world changes.
For this reason, we require the world-indexed memory invariants to be monotone in the world.
This means that a pair of memory segments that are related now will stay related in future worlds.
This changes the spatial regions as follows:
\[
  \Regions = \left\{
  \begin{array}{l}
    \{\spatial \} \times (\World \monfun \Rel{\MemSeg \times \MemSeg}) + \\
    \{\spatialo \} \times (\World \monfun \Rel{\MemSeg \times \MemSeg}) + \\
    \{\revoked \}
  \end{array} \right.
\]
We make a similar change to $\Regionh$.
% TODO update terminology: memory invariant, H, and seal invariant, H_\sigma

\subsubsection{Solving the recursive domain equation}
\label{subsubsec:rec-dom-eq}
% TODO consider new structure (the below)
% First introduce c.o.f.e.s and the structures you can freely use to get a solution for your recursive domain equation
% Then introduce all the specialized gunk necessary for our worlds
%
%solving the recursive domain equation means that we can present the actual definitions
% Also see the notes from Aina
In the previous sections, we sketched what our worlds should be.
However, the worlds we want constitute a self-referential domain equation for which no solution exists in set theory.
% Move to a different domain
Therefore, we need to move to a different domain with enough structure for a solution to exist for recursive equations.
We move to a setting where instead of sets we have c.o.f.e.'s (complete ordered families of equivalences), instead of functions we have non-expansive functions, and instead of relations we have downwards-closed relations.
Intuitively, a c.o.f.e.\ is a set with some extra structure.
Specifically, c.o.f.e.'s have a step-index that must satisfy a set of conditions which provide sufficient structure for a solution to exist for the recursive domain equation.
In this section we present the relevant definitions, the final world domain equation, and sketch what needs to be proven to get a solution to the recursive domain equation.

% Cofe's
In the following, we present complete ordered family of equivalences \citep{di_gianantonio_2002}.
We loosely follow the presentation in \citet{birkedal_taste_2014} and introduce the necessary definitions needed for our particular world.
\begin{definition}[Ordered family of equivalences (o.f.e.)]
An \emph{ordered family of equivalences} is a pair $\cofe{X}$ where $X$ is a set and $\nequal$ is an equivalence relation for each $n$.
The pair must satisfy the following properties:
\begin{enumerate}
\item $\nequal[0]$ is the total relation.
\item For all $n \in \nats$, $\nequal[n] \supseteq \nequal[n+1]$.
\item For all $x,y \in X$,  if for all $n \in \nats$ $x \nequal y$, then $x = y$.
\end{enumerate}
\end{definition}
\noindent The index on the family of equivalences is intuitively a measure of precision.
The larger the index is, the more refined the equivalences are which means that they can distinguish more elements from one another.
On the other hand as the index decreases, the equivalences become increasingly imprecise and unable to distinguish certain elements.
At index 0, all precision is lost and the equivalence cannot distinguish anything.
% TODO: Lau write intuition for sequences and why they are necessary.
\begin{definition}[Complete ordered family of equivalences (c.o.f.e.)]
  \label{def:cauchy-sequence}
  A \emph{complete ordered family of equivalences} is an ordered family of equivalences
  $\cofe{X}$ such that every Cauchy sequence in $X$ has a limit
  in $X$.

@@

  An element $x \in X$ is the \emph{limit} of the sequence $\seq{x}$ if
  \begin{align*}
    \forall k \in \nats, \exists j \in \nats, \forall n \geq j, x \nequal[k] x_n.
  \end{align*}
\end{definition}
% TODO write something here about why the different things (the meassure of the ofe and the guaranteed limits of the cofe) gives us in terms of finding a solution.
In order to move from one c.o.f.e.\ to another, we define functions, just like we would when working with sets.
Unlike functions between sets, functions between c.o.f.e.'s must preserve the c.o.f.e.\ structure which informally means that the measure of precision must be retained.
This specifically means that functions between c.o.f.e.'s must be \emph{non-expansive} which basically means that they preserve $n$-equalities.
Functions may not only retain equivalences but make it more precise.
Such a function is said to be \emph{contractive}.
\begin{definition}
  \label{def:nonexpansive-contractive-ofe}
  Let $\left(X,\left(\nequal[n]_X\right)_{n=0}^{\infty} \right)$ and $\left(Y,\left(\nequal[n]_Y\right)_{n=0}^{\infty} \right)$ be two ordered families of equivalences and $f$ a function from the set $X$ to the set $Y$.
  We 
  The function $f$ is
  \begin{description}
  \item[non-expansive] when for all $x, x' \in X$, and all $n \in \nats$,
\[
  x \nequal_X x' \implies f(x) \nequal_Y f(x')
\]
  \item[contractive] when for any $x, x' \in X$, and any $n \in \nats$,
\[
  x \nequal_X x' \implies f(x) \nequal[n+1]_Y f(x') 
  \qedhere
\]
  \end{description}
\end{definition}
\noindent The step-indexing provided by c.o.f.e.'s give enough structure to solve the recursive domain equation.
There are a number of standard c.o.f.e.\ constructions for countably infinite sets, products, sums, and functions.
A set can be lifted to c.o.f.e.\ by having the normal equality for all positive indices and the total relation for $\nequal[0]$.
The other constructions are used to create c.o.f.e.'s from other c.o.f.e.'s.
For instance the product construction is defined as follows:
\begin{definition}[Product c.o.f.e.]
\label{def:product-cofe}
Given two c.o.f.e.'s $\left(X, \left(\nequal_X\right)_{n=0}^{\infty} \right)$ and $\left(Y, \left(\nequal_Y\right)_{n=0}^{\infty} \right)$ define the product c.o.f.e.\ as $\left(X \times Y, \left(\nequal \right)_{n=0}^{\infty} \right)$ where the equivalence family is defined as for $(x,y),(x',y') \in X \times Y$
\[
  (x,y) \nequal (x',y') \text{ iff } x \nequal x' \wedge y \nequal y'
\]
\end{definition}
The remaining standard constructions are defined similarly.
Another important construction is the \emph{later} ($\blater$) c.o.f.e.\
The later c.o.f.e.\ construction is based on one c.o.f.e.\ which it decreases the index of.
\begin{lemma}[$\blater$ c.o.f.e.]
  \label{def:later-cofe}
  Given a c.o.f.e $\left(X, \left(\nequal_X\right)_{n=0}^{\infty} \right)$ define the later c.o.f.e.\ as
  \[
    \blater\left(X, \left(\nequal_X\right)_{n=0}^{\infty} \right) = \left(X, \left(\nequal[n]_{\blater}\right)_{n=0}^{\infty} \right)
  \]
  where
  \[
    x \nequal_{\blater} x' \text{ iff } \left\{
      \begin{array}{l}
        n = 0 \vee{}\\
        x \nequal[n+1]_X x'
      \end{array}
    \right.
  \]
\end{lemma}
The later c.o.f.e.\ is used to make sure that the index decreases before the self-reference which gives something to recurse on in order to find a solution for the domain equation.
If we only use the standard constructions to construct a self-referential equation and make sure to guard every negative occurrence of the self reference with the later c.o.f.e.\ construction, then the equation has a solution.
\todo[inline]{Insert the solution theorem here?}

@@

>>>>

We call the invariants regions because, as we will see in Section~\ref{subsubsec:ft-and-revocation}, they turn out to not be invariant.
A region describes a collection of related memory segments, so it is simply represented as a relation over memory segments (i.e.\ a relation in $\Rel{\MemSeg \times \MemSeg}$).
Intuitively, we want to be able to say that two memory segments are related when their contents are related.
That is, for every address in the two memory segments, the words that reside there must be related.
In Section~\ref{subsec:logical-relation}, we define precisely what it means for words to be related, but for now we provide some intuition.
If we have two integers, then they are related when they are equal.
If we instead have two capabilities, then they should also be related if they, in some sense, are equal.
But what does it mean for capabilities to be \textit{equal}?
Intuitively, it should mean that the capabilities give you the same authority, for instance, if you have two executable capabilities, they should observably perform the same computation.
As a capability points to a piece of memory, the authority of the capability depends on the contents of that memory.
In other words to say whether two capabilities are related, we must know the possible contents of the memory.
The world expresses the possible memory contents, so our regions are world indexed, i.e.
\[
  \Worldh = \RegName \parfun (\World \fun \Rel{\MemSeg \times \MemSeg})
\]
At this point, we can see that we have constructed a recursive domain equation.
If we inline $\Worldh$ in $\World$, then we have a circular equation with no solution because the self-reference happens in a negative position.
Luckily, we can solve circular equations if we move to a different domain.
For now, we will ignore the problem and return to the issue in Section~\ref{subsubsec:rec-dom-eq}.

For the sake of readability, we introduce the following notation
\begin{gather*}
  \pwheap = \pi_1(W)\\
  \pwpriv = \pi_2(W)\\
  \pwfree = \pi_3(W)
\end{gather*}
\subsubsection{Linearity}
% Linearity
The linear capabilities of \srccm{} and \trgcm{} guarantee that they have the sole authority over the memory they reference\footnote{Under the assumption that the system was initialised with unique linear capabilties.}.
In order to model this uniqueness, we need to keep track of which parts of memory are uniquely referenced and make sure that only one linear capability references the unique parts.
%% Bookkeeping done by worlds
We use the world to keep track of what parts of memory must be uniquely referenced by having two kinds of regions: shared and spatial.
If a memory segment is governed by a shared region, then normal capabilities may reference it.
On the other hand, if a memory segment is governed by a spatial region, then only a linear capability may reference it.
%%% spatial/spatial_owned
\todo[inline]{Dominique: Maybe it would be clearer to rename spatial\_owned/spatial to spatial/spatial\_shadow or something similar. Spatial\_owned is really the default meaning of spatial, and what we've called spatial is a peculiar thing that we use for a very specific purpose.
  This would make the discussion easier to follow, I think.}

We cannot let multiple linear capabilities reference the same memory, so we add ownership to spatial regions.
Only the spatial owned regions can be referenced by a capability.
Even though only spatial owned regions can be referenced by linear capabilities, we still have unclaimed spatial regions that specify what part of memory looks like.
When there is an unclaimed spatial region, then there may be a compatible world  that claims the ownership of that spatial region which means that the memory may still be referenced (we return to the notion of compatible worlds w.r.t. ownership in Section~\ref{subsubsec:joining-worlds}).
Specifically, we add tags $\spatial$ and $\spatialo$ to the spatial regions:
\[
  \Regions = \left\{
  \begin{array}{l}
    \{\spatial \} \times (\World \fun \Rel{\MemSeg \times \MemSeg}) + \\
    \{\spatialo \} \times (\World \fun \Rel{\MemSeg \times \MemSeg}) \\ 
  \end{array} \right.
\]
\todo[inline]{Dominique: Perhaps rename pure to shared?  Or introduce the term pure earlier on?}
For readability, we also add a tag $\pure$ to the shared regions:
\[
  \Regionh = \{\pure \} \times (\World \fun \Rel{\MemSeg \times \MemSeg}) 
\]

% What parts of memory are linear
We will extend the regions further in Sections~\ref{subsubsec:seals}~and~\ref{subsubsec:ft-and-revocation}, but for now we continue the definitions of the three sub-worlds.
The sub-world $\Worldh$ specifies the heap memory which can be referenced by both linear and normal capabilities, so it should contain both shared and spatial regions.
For this reason, it is defined as
\[
  \Worldh = \RegName \parfun (\Regionh + \Regions)
\]
\todo[inline]{Dominique: $+$?  Why not $\cup$?}
\srccm{} internalizes the \stktokens{} stack, so it can only be referenced by linear capabilities which means that the two stack regions should only have spatial regions.
For instance the $\Worldfs$ is defined as
\[
  \Worldfs = \RegName \parfun \Regions
\]
$\Worlds$ can also only be referenced by linear capabilities, so it should also only have spatial regions.
$\Worlds$ not only models the memory contents of the local stack frames, it also models the call stack that consists of the stack frames.
Conceptually, this means that each of the stack frames is connected with a return point in some code.
In a traditional C calling convention, the code return point would even be stored in the stack frame.
However, with \stktokens{} a stack frame does not contain any information about the corresponding code return point.
Instead, the stack frame and code return point is connected by the capabilities that reference them as they are sealed with the same seal and together then constitute the sealed return pair.
In order to ensure that each call actually returns to the correct point of the code, we must still include the address of the return point in our model.
To this end, each shared region in $\Worlds$ must be paired with a return address:
\[
\Worlds = \RegName \parfun (\Regions \times \Addr)
\]

@@

\subsubsection{Future worlds and revocation}
\label{subsubsec:ft-and-revocation}
% Memory evolves over time, we need to model this.
During machine execution, the memory contents changes and parts of memory may be repurposed\todo[inline]{Dominique: Add something like ``, at least when we can be sure that there are no capabilities left in the system whose validity depends on the old contents of the memory.''?}.
The world must model such memory changes.
% Future World
To this end, we have a future world relation $\future$ that describes how worlds and thus memory may change over time.
\todo[inline]{Rather than introducing the future world relation as ``how memory may change over time'', then its precise definition and then monotonicity, I would explain monotonicity first and the precise definition of the future world relation second.
  IMHO, it makes a lot more sense if you think about the future world as modelling precisely those changes in assumptions that do not invalidate the relatedness of values.
  In other words, I would say something like ``Very often, it is the case that relatedness of two capabilities doesn't change if extra assumptions in the system are added. For example, two related capabilities remain related when an extra invariant is added on unrelated memory, or when a stack frame that it does not reference is dropped.
  In Kripke logical relations, this kind of assumption changes, that do not invalidate the relatedness of values, are modelled by the future world relation.
  etc.''
  }

  \todo[inline]{Maybe explain here that Kripke future world relations usually only allow for adding fresh assumptions on fresh memory or protocol steps that the system was designed to support from the start.
    However in our setting, we sometimes allow dropping assumptions, namely when linearity tells us that no value in the system depends on the capability anymore.
    By linking revocation of regions immediately to linearity, it will require less explanation later that only spatial regions can be revoked.
  }
In order to relate \trgcm{} and \srccm{}, we are mainly interested in changes to the stack as it repurposes stack frames.
Specifically, we need to introduce regions to govern new encapsulated stack frames, and we need to revoke regions for call frames that are popped from the call stack, to make place for a new frame.
We allow introduction of new regions simply by extension.
That is, a future world can have new regions that was not present in a past world.
Things are a bit more complicated when it comes to revocation.
\lau{Can we give a more precise explanation of why regions cannot be removed from the world (still in terms of intuition).
}
\dominique{To be honest, I don't actually know what exactly would go wrong if we simply removed revoked regions.
  Perhaps we should think this through?
}
For technical reasons, we need the worlds to be extensional\todo{Dominique: meaning?}, so we cannot simply forget about regions that we want to remove.
We keep removed regions in the world and tag them as removed with a $\revoked$ tag.
This means that the regions should be able to get the $\revoked$ tag which we express in a future region relation.

Not all regions should be revocable.
\todo[inline]{
  Perhaps explain intuition with a phrase like ``Since regions express assumptions that the validity of other values in the system may depend on, we should only be able to revoke a region when we are sure that nothing depends on it.
  This is never the case for shared regions, as blabla copying blabla...
  However, if we have a linear capability to a block of memory, then linearity ensures that no other capability in the system can reference the same memory.''
  }

Specifically, it should only be the regions that specify memory governed by linear capabilities as linear capabilities, in some sense, can be revoked.
A linear capability is the sole authority over some memory, so when a program possesses a linear capability, it knows for sure that no one else has access to that memory which makes it safe to repurpose.
On the other hand, normal capabilities may have an alias, so it is not safe to repurpose the memory a normal capability has authority over.
Only spatial regions specify memory governed by linear capabilities, so we only add the revoked tag to the $\Regions$.
The $\revoked$ region does not specify any memory, so we drop the world indexed relation.
That is, we have
\[
  \Regions = \left\{
  \begin{array}{l}
    \{\spatial \} \times (\World \fun \Rel{\MemSeg \times \MemSeg}) + \\
    \{\spatialo \} \times (\World \fun \Rel{\MemSeg \times \MemSeg}) + \\
    \{\revoked \}
  \end{array} \right.
\]
The rules for the future region relation are displayed in Figure~\ref{fig:ft-reg-rel}.
% Revocation of spatial regions. 'revocation' of linear capabilities.
The first rule says that all regions are future regions of themselves, so regions are allowed to stay unchanged.
The second rule says that $\spatial$ regions can become $\revoked$ which means that nothing can depends on the region anymore.
The third and final rule says that $\spatial$ regions can become $\spatialo$ which indicates that the linear capabilities are affine.
Specifically, the linear capabilities in \trgcm{} and \srccm{} can be dropped by overwriting them in registers or memory which makes the linear capabilities affine.
The world a capability is safe with respect to is no longer needed when the capability is dropped, so other worlds may pick up the ownership which the future region relation allows.
% describe why "Spatial -> spatialo" ~ affine or at least why we have it?
\begin{figure}[htb]
  \centering
  \begin{mathpar}
  \inferrule{ r \in \Regions \cup \Regionh }{ r \future r}
  \and
  \inferrule{ }{ \revoked \future (\spatial,\_)}
  \and
  \inferrule{ }{ (\spatialo,H) \future (\spatial,H)}
\end{mathpar}
  \caption{Future region relation.}
  \label{fig:ft-reg-rel}
\end{figure}

\todo[inline]{Dominique: it would be great if we could add an example here of two concrete capabilities, one of which is linear, the worlds in which they're valid and explain how we can revoke the linear one and repurpose the memory it addresses, but still keep the other capability valid in the new world?
}

With the future region relation in place, we define the future world relation as follows: For worlds $W$ and $W'$,
\[
  W' \future W \text{ iff } \left\{
    \array{l}
    \text{for $i \in \{\mathrm{heap},\mathrm{free},\mathrm{priv} \}$} \\
    \quad\exists m_i : \RegionName \fun \RegionName, \text{ injective}\ldotp\\
    \qquad \dom(W'.i) \supseteq \dom(m_i(W.i)) \wedge \forall r \in \dom(W.i)\ldotp W'.i(m_i(r)) \future W.i(r)
     \endarray
  \right.
\]
The relation says that each of the three worlds must be an extension of the past world and each of the existing regions must have a future region.
% injective function
Note that the future world relation has a mapping function $m_i$ which allows us to change the naming of regions in future worlds.
The definition is a generalization of the standard definition where $m_i$ would be the identity\footnote{In \citet{skorstengaard_reasoning_2017} the future region relation and the reasoning about the awkward example could have been simplified with this future world relation.}.\dominique{I wonder if the renaming function obviates the need for the ``revoked'' trick.}

The memory invariants are world dependent, but the worlds may change according to the future world relation, so what does that mean for the memory invariants?
The memory invariants describe what certain parts of memory looks like.
In future worlds, part of the memory may change but for the memory invariant to be invariant, it should continue relating the same memory segments even when the rest of the world changes.
\dominique{This discussion might be easier if we had talked about monotonicity earlier.}
For this reason, we require the world-indexed memory invariants to be monotone in the world.
This means that a pair of memory segments that are related now will stay related in future worlds.
This changes the spatial regions as follows:
\[
  \Regions = \left\{
  \begin{array}{l}
    \{\spatial \} \times (\World \monfun \Rel{\MemSeg \times \MemSeg}) + \\
    \{\spatialo \} \times (\World \monfun \Rel{\MemSeg \times \MemSeg}) + \\
    \{\revoked \}
  \end{array} \right.
\]
We make a similar change to $\Regionh$.
% TODO update terminology: memory invariant, H, and seal invariant, H_\sigma

\subsubsection{Solving the recursive domain equation}
\label{subsubsec:rec-dom-eq}
\dominique{Explain that this section provides technical details on how the recursive world equation can be solved and that readers may skip this if they like.}
% TODO consider new structure (the below)
% First introduce c.o.f.e.s and the structures you can freely use to get a solution for your recursive domain equation
% Then introduce all the specialized gunk necessary for our worlds
%
%solving the recursive domain equation means that we can present the actual definitions
% Also see the notes from Aina
In the previous sections, we sketched what our worlds should be.
However, the worlds we want constitute a self-referential domain equation for which no solution exists in set theory.
% Move to a different domain
Therefore, we need to move to a different domain with enough structure for a solution to exist for recursive equations.
\dominique{mention that this is a standard technique (cite the relevant papers) and an abstracted form of step-indexing (cite some more papers).}
We move to a setting where instead of sets we have c.o.f.e.'s (complete ordered families of equivalences), instead of functions we have non-expansive functions, and instead of relations we have downwards-closed relations.
Intuitively, a c.o.f.e.\ is a set with some extra structure.
Specifically, c.o.f.e.'s have a step-index that must satisfy a set of conditions which provide sufficient structure for a solution to exist for the recursive domain equation.
In this section we present the relevant definitions, the final world domain equation, and sketch what needs to be proven to get a solution to the recursive domain equation.

% Cofe's
In the following, we present complete ordered family of equivalences \citep{di_gianantonio_2002}.
We loosely follow the presentation in \citet{birkedal_taste_2014} and introduce the necessary definitions needed for our particular world.
\begin{definition}[Ordered family of equivalences (o.f.e.)]
An \emph{ordered family of equivalences} is a pair $\cofe{X}$ where $X$ is a set and $\nequal$ is an equivalence relation for each $n$.
The pair must satisfy the following properties:
\begin{enumerate}
\item $\nequal[0]$ is the total relation.
\item For all $n \in \nats$, $\nequal[n] \supseteq \nequal[n+1]$.
\item For all $x,y \in X$,  if for all $n \in \nats$ $x \nequal y$, then $x = y$.
\end{enumerate}
\end{definition}
\dominique{Explain how an o.f.e. can be seen as a set with extra structure.}
\noindent The index on the family of equivalences is intuitively a measure of precision.
\dominique{Introduce phrases like ``up to n steps'' etc.}
The larger the index is, the more refined the equivalences are which means that they can distinguish more elements from one another.
On the other hand as the index decreases, the equivalences become increasingly imprecise and unable to distinguish certain elements.
At index 0, all precision is lost and the equivalence cannot distinguish anything.
% TODO: Lau write intuition for sequences and why they are necessary.
\begin{definition}[Complete ordered family of equivalences (c.o.f.e.)]
  \label{def:cauchy-sequence}
  A \emph{complete ordered family of equivalences} is an ordered family of equivalences
  $\cofe{X}$ such that every Cauchy sequence in $X$ has a limit
  in $X$.

@@

  An element $x \in X$ is the \emph{limit} of the sequence $\seq{x}$ if
  \begin{align*}
    \forall k \in \nats, \exists j \in \nats, \forall n \geq j, x \nequal[k] x_n.
  \end{align*}
\end{definition}
% TODO write something here about why the different things (the meassure of the ofe and the guaranteed limits of the cofe) gives us in terms of finding a solution.
In order to move from one c.o.f.e.\ to another, we define functions, just like we would when working with sets.
Unlike functions between sets, functions between c.o.f.e.'s must preserve the c.o.f.e.\ structure which informally means that the measure of precision must be retained.
This specifically means that functions between c.o.f.e.'s must be \emph{non-expansive} which basically means that they preserve $n$-equalities.
Functions may not only retain equivalences but make it more precise.
Such a function is said to be \emph{contractive}.
\begin{definition}
  \label{def:nonexpansive-contractive-ofe}
  Let $\left(X,\left(\nequal[n]_X\right)_{n=0}^{\infty} \right)$ and $\left(Y,\left(\nequal[n]_Y\right)_{n=0}^{\infty} \right)$ be two ordered families of equivalences and $f$ a function from the set $X$ to the set $Y$.
  The function $f$ is
  \begin{description}
  \item[non-expansive] when for all $x, x' \in X$, and all $n \in \nats$,
\[
  x \nequal_X x' \implies f(x) \nequal_Y f(x')
\]
  \item[contractive] when for any $x, x' \in X$, and any $n \in \nats$,
\[
  x \nequal_X x' \implies f(x) \nequal[n+1]_Y f(x') 
  \qedhere
\]
  \end{description}
\end{definition}
\noindent The step-indexing provided by c.o.f.e.'s give enough structure to solve the recursive domain equation.
There are a number of standard c.o.f.e.\ constructions for countably infinite sets, products, sums, and functions.
A set can be lifted to c.o.f.e.\ by having the normal equality for all positive indices and the total relation for $\nequal[0]$.
The other constructions are used to create c.o.f.e.'s from other c.o.f.e.'s.
For instance the product construction is defined as follows:
\begin{definition}[Product c.o.f.e.]
\label{def:product-cofe}
Given two c.o.f.e.'s $\left(X, \left(\nequal_X\right)_{n=0}^{\infty} \right)$ and $\left(Y, \left(\nequal_Y\right)_{n=0}^{\infty} \right)$ define the product c.o.f.e.\ as $\left(X \times Y, \left(\nequal \right)_{n=0}^{\infty} \right)$ where the equivalence family is defined as for $(x,y),(x',y') \in X \times Y$
\[
  (x,y) \nequal (x',y') \text{ iff } x \nequal x' \wedge y \nequal y'
\]
\end{definition}
The remaining standard constructions are defined similarly.
Another important construction is the \emph{later} ($\blater$) c.o.f.e.\
The later c.o.f.e.\ construction is based on one c.o.f.e.\ which it decreases the index of.
\begin{lemma}[$\blater$ c.o.f.e.]
  \label{def:later-cofe}
  Given a c.o.f.e $\left(X, \left(\nequal_X\right)_{n=0}^{\infty} \right)$ define the later c.o.f.e.\ as
  \[
    \blater\left(X, \left(\nequal_X\right)_{n=0}^{\infty} \right) = \left(X, \left(\nequal[n]_{\blater}\right)_{n=0}^{\infty} \right)
  \]
  where
  \[
    x \nequal_{\blater} x' \text{ iff } \left\{
      \begin{array}{l}
        n = 0 \vee{}\\
        x \nequal[n+1]_X x'
      \end{array}
    \right.
  \]
\end{lemma}
The later c.o.f.e.\ is used to make sure that the index decreases before the self-reference which gives something to recurse on in order to find a solution for the domain equation.
If we only use the standard constructions to construct a self-referential equation and make sure to guard every negative occurrence of the self reference with the later c.o.f.e.\ construction, then the equation has a solution.
\todo[inline]{Insert the solution theorem here?}

@@

\dominique{At this point, the paper is starting to read like a tech report.
  Are we sure it's a good idea to provide this much technical detail?
  I would propose to not include all definitions, but a representative selection and refer to the TR for full detail.
}
